---
title: "Course technology overview"
editor: visual
code-copy: true
execute:
  message: false
  warning: false
  echo: true
  cache: true
---

Read this and complete all instructions in the 'action' boxes during your lab section. Your TA will walk you through the activity and help to troubleshoot issues and answer any questions along the way.

**Objectives:**

-   get oriented with RStudio and RStudio projects

-   install packages

-   accept a GitHub classroom team assignment

-   clone a repository

## Prerequisites

<!--# TA ACTION: ensure everyone has working installs and a github account before going ahead. there may be time to allow a few minutes for anyone who needs the time to complete installs, but if need be you can pair students with technical difficulties with students with working setups and troubleshoot after class. -->

To complete the activity you'll need to:

-   have all of the software listed on the course technology page installed;

-   find (or create) your [GitHub](https://github.com/) account credentials (if you are creating an account for the first time, see [advice on choosing a username](https://happygitwithr.com/github-acct.html#username-advice)).

::: callout-important
## Action

**Preparations:**

1.  Log in to your GitHub account.
2.  Open your GitHub client.
3.  Open a new session in RStudio.
4.  Create a class folder for PSTAT197A somewhere on your machine, *e.g.,* `~/documents/pstat197a`.
:::

## RStudio projects

First we'll get acquainted with the basic functionality of the RStudio IDE and the use of projects as a means of organizing files. If you've already used RStudio, great -- this will still serve to introduce you to how we'll use RStudio projects in this class.

### RStudio Setup

<!--# TA ACTION: take a minute or two to review the IDE layout in RStudio. point out the bulletted items and say a word or two about each one. -->

Your TA will briefly review the (default) layout of the RStudio IDE. You should be able to identify/find the following:

-   console

-   terminal

-   file navigator

-   environment

-   history

We'll use several R packages throughout the quarter. Some of these we will install on the go, but we can install several that we'll rely on now.

::: callout-important
## Action

**Install packages**

Navigate to the console and copy-paste the following commands. You only need to do this once. This will take a minute or two to complete.
:::

```{r, eval=F}
# package install list 
url <- 'https://raw.githubusercontent.com/pstat197/pstat197a/main/materials/scripts/package-installs.R'
source(url)

# clear environment
rm(list = ls())
```

### Create a local project

*Projects* are a means of keeping your work organized. When you create a project in a directory on your local machine, RStudio keeps track of project metadata, history, and the working environment so that every time you open the project you see whatever you had open when you last closed it.

::: callout-important
## Action

**Create a new project**:

-   Select *File \> New project*

-   Create the project in a new directory as a subdirectory of your class folder

-   Name it `example-project`

*Comment:* when naming files it's good practice to avoid spaces, special characters, and the like. A naming convention we try to follow: choose a descriptive name comprising 1-3 words or common abbreviations separated by hyphens.
:::

Take a moment to observe the file navigator. It should consist of a single `example-project.Rproj` file.

### Add content

We may as well populate the project with a few files -- so let's add a dataset and write a short script, as if we're just starting a data analysis.

::: callout-important
## Action

**Retrieve data and store a local copy**

-   Open a new script: *File \> New File \> R Script*

-   In the navigator, create a folder called `data` and a folder called `scripts`

-   Copy and paste the code chunk below into your script.

-   Execute once, then save in the `scripts` folder as `data-retrieval.R` and close
:::

```{r, eval = F}
library(tidyverse)

# retrieve pollution data
url <- 'https://raw.githubusercontent.com/pstat197/pstat197a/main/materials/labs/lab1-setup/data/pollution.csv'
pollution <- read_csv(url)

# write as csv to file
write_csv(pollution, file = 'data/pollution.csv')

# clear environment
rm(list = ls())
```

<!--# TA COMMENT: you might pause here and mention a few things. first, we've used a script (not an rmarkdown file) because we only have R codes and no text content (many students acquire the bad habit of working in rmarkdown when writing codes). second, storing a script like this might seem silly but it does establish a record of the data source -->

Next, we'll do a simple regression analysis.

::: callout-important
## Action

**Create a script**

-   Create a new script as before

-   Copy-paste the code chunk below into your script

-   Execute once and examine the results

-   Save in the `scripts` folder as `slr-analysis.R`
:::

```{r, message = F, warning = F}
library(tidyverse)

# load data
pollution <- read_csv('data/pollution.csv')

# examine scatterplot with SLR fit
ggplot(pollution,
       aes(x = log(SO2), y = Mort)) +
  geom_point() +
  geom_smooth(method = 'lm')

# compute SLR fit
fit <- lm(Mort ~ log(SO2), data = pollution)
broom::tidy(fit)

# interpret
fit_ci <- confint(fit, parm = 'log(SO2)')*log(1.2)

paste('With 95% confidence, every 20% increase in sulfur dioxide pollution is associated with an increase in two-year mortality rate between', 
      round(fit_ci[1], 2), 
      'and', 
      round(fit_ci[2], 2), 
      'per 100k', sep = ' ') %>% 
  print()
```

<!--# TA COMMENT: show everyone what their directories should look like at completion. mention that the project treats the project directory as the working directory, so scripts stored anywhere in the project still parse directory paths relative to the project directory, NOT the file location. -->

Congrats on your first project! You can close the RStudio session now.

We'll be using projects structured much like what you just set up, but with one catch: we'll link up our RStudio projects with shared *repositories* so that we can all collaborate on the same set of project files.

## GitHub repositories

<!--# TA COMMENT: you could either talk through this section or have students read it and clone the repo themselves. either way, show the use of the client, the local copy, and the remote location (on github.com). the goal here is for students to understand the distinction between local and remote and that the client provides a means of communication -->

We will be using *repositories* on GitHub throughout the course. A repository is simply a storage space for the files associated with a project.

For the activities in PSTAT 197A, you will work with several different groups over the course of the quarter. Each group will use a shared GitHub repository for the module you are completing together.

All course repositories should be created within the **PSTAT 197 course GitHub organization** rather than under an individual student's personal GitHub account.

Before completing this activity, you should have received and accepted an invitation to join the course GitHub organization.

For now, your group will create a **group sandbox repository** that you can use to practice working collaboratively with Git and GitHub during our next class meeting.

::: callout-important
## Action

**Create your group's sandbox repository**

1. Go to the PSTAT 197 course organization on GitHub.
2. One person at your table should create a new repository in the organization.
3. Name the repository according to the naming convention provided by your instructor, such as `sandbox-group-01`.
4. Set the repository to **Public**.
5. Do not initialize it with any additional files unless instructed to do so.
6. After creating the repository, add the other members of your table group as collaborators with permission to contribute to the repository.
7. Each group member should confirm that they can access the repository on github.com.

Keep the repository page open; you will need its URL shortly.
:::

<!--# TA COMMENT: make sure only ONE student per group creates the repository. The other students should be added to that repository rather than creating additional copies. -->

### Git and GitHub

At some point in time -- possibly quite recently -- you had to install Git on your local machine, as well as create a GitHub account. Git and GitHub are two different things.

**Git is version control software** that enables you to systematically track and control file changes within a *repository* -- a collection of files, possibly with some directory structure. (The definition of 'repository' is simply 'storage place'.)

**GitHub is an online platform** for hosting repositories remotely. A public repository can be viewed by anyone, but only people with appropriate permissions can directly contribute changes to it.

This allows multiple members of a group to collaborate on the same collection of files while Git keeps track of the changes that are made.

### local <> remote

Usually remote repositories are not updated by directly editing files on github.com. Contributors generally need to execute code and test their changes on their own computers before sharing those changes with the rest of the group.

Instead, contributors prepare changes on their own machines, where they can run and test their code, and then update the remote repository once their changes are ready.

This process involves communicating information between *local* and *remote* locations. For this purpose, each contributor needs a local copy of the remote repository.

### Cloning a repository

In Git terminology, a *clone* is a local copy of a remote repository. Creating a clone copies the repository files to your computer and establishes the connection between the local and remote repositories so that changes can be sent to and received from GitHub.

***You only need to create a clone once for each repository.***

To clone a repository, all one needs is:

- the remote repository URL;
- the local destination where the clone will be created;
- permission to access the repository, if the repository is private.

Here you'll clone the group sandbox repository your group just created. You will need its URL. If you happened to close the page, you can find the repository by returning to the PSTAT 197 organization on github.com.

<!--# TA COMMENT: show students how to navigate to the organization and find their group's repository if necessary. Emphasize that every member of the group clones the SAME remote repository. -->

::: callout-important
## Action

**Clone the sandbox repository**

1. Open your GitHub client (GitKraken, GitHub Desktop, or similar) and ensure you are logged in to your GitHub account.
2. Find the option to clone a repository.
3. Select or enter the URL of your group's sandbox repository.
4. Choose the location on your computer where you would like the local copy to be stored.
5. Complete the cloning process.
6. Check your file navigator to confirm that the repository was copied to your computer.

Every member of the group should clone the same repository onto their own computer.
:::

<!--# TA COMMENT: remind students we'll use this repository next class meeting, so DON'T delete the clone -->

An alternative is to create the clone using a terminal command. In the terminal, navigate to the desired destination and enter:

```bash
git clone https://github.com/ORGANIZATION/REPONAME
```

For example:

```bash
git clone https://github.com/pstat197-2026-27/sandbox-group-01
```

::: callout-note
## Remarks

**On terminal commands:**

- It's recommended to manage Git actions through a visual client at first, as it's much easier to see and understand what's happening.
- However, if you know exactly what you're doing, executing simple actions via Git in the terminal can be more efficient at times.
- For example, you can keep a terminal open in RStudio and manage your repository workflow from there, without having to toggle between environments.
- Try experimenting with terminal commands from RStudio after you have a little experience with basic Git actions.
:::

### Repositories for course modules

You will follow essentially the same process for each of the five PSTAT 197A modules.

For each module:

1. You will be assigned to a group for that module.
2. One member of the group will create a new **public repository** within the PSTAT 197 organization.
3. That student will add the other members of the group as collaborators.
4. Every group member will clone the same repository to their own computer.
5. The group will use that repository for its work on that module.

Your groups may change from one module to another. A new repository should therefore be created for each module rather than continuing to use a repository belonging to a previous group.

These module groups are for practicing the skills you will need later in the capstone sequence; they are **not your final capstone project groups**.

## Checklist

Have you completed all of the activity action items?

1. Install software: R, RStudio, Git, and a GitHub client
2. Create a GitHub account
3. Accept the invitation to join the PSTAT 197 GitHub organization
4. Install R packages that will be used frequently
5. Create a local project in RStudio
6. Create or join your group's sandbox repository
7. Clone the group sandbox repository