---
title: "Tidyverse basics"
editor: visual
code-copy: true
execute:
  message: false
  warning: false
  echo: true
  cache: false
---

Read through the R Basics section and then complete all actions in the Tidyverse basics section. This lab is for your own benefit and no submission is expected.

**Objectives:**

-   Review R basics (data types and object classes);

-   introduce core tidyverse libraries (dplyr, tidyr, and ggplot);

-   replicate portions of in-class analysis of class survey data

# R basics for total beginners

Here we'll cover some bare essentials in R. You can find a more thorough introduction in [MSDR Appendix B](https://mdsr-book.github.io/mdsr2e/ch-R.html#ch:R).

## Data types

We will focus on four common atomic data types in R and one important class for categorical data: factors.

***Numeric*** data represent real numbers and are normally stored as double-precision floating-point values in R. Numeric data are abbreviated `num` and by default are stored as double-precision floating point.

```{r}
# a number
4.5

# check structure
str(4.5)

# stored as double
is.double(4.5)
```

***Integer*** data are integers. For the most part they behave like numeric data, except occupy less memory, which can in some cases be convenient. To distinguish integers from doubles, R uses a trailing `L` after values; the data type is abbreviated `int`.

```{r}
# an integer
4L

# check structure
str(4L)
```

***Logical*** data are binary and represented in R as having values `TRUE` and `FALSE`. They are abbreviated `logi` in R. Often they are automatically coerced to integer data with values 0 (false) and 1 (true) to perform arithmetic and other operations.

```{r}
# logical value
TRUE

# check structure
str(TRUE)

# arithmetic
TRUE + FALSE

# check structure
str(FALSE + FALSE)
```

***Character*** data represent strings of text and are sometimes called 'strings'. They are abbreviated `chr` in R and values are surrounded by quotation marks; this distinguishes, for example, the *character* 4 from the *number* 4. Single quotations can be used to input strings as well as double quotations. Arithmetic is not possible with strings for obvious reasons.

```{r}
#| error: true

# a character string
'yay'

# check structure
str('yay')

# string arithmetic won't work
# '4' + '1'

# but can be performed after converting character strings to numbers
as.numeric('4') + as.numeric('1')
```

***Factors*** represent categorical variables. A factor is an object class built on an integer vector, rather than a separate atomic data type. In R these are encoded numerically according to the number of 'levels' of the factor, which represent the unique values of the categorical variable, and each level is labeled. R will print the labels, not the levels, of factors; the data type is abbreviated `fct`.

```{r}
# a factor
factor(1, levels = c(1, 2), labels = c('blue', 'red'))

# less verbose definition
factor('blue', levels = c('blue', 'red'))

# check structure
str(factor('blue', levels = c('blue', 'red')))
```

CSV files do not preserve factor definitions. In this lab, `read_csv()` imports proficiency labels as character data, and we explicitly convert them to ordered factors. The levels and labels of factors can be manipulated using a variety of helper functions.

## Object classes

The most basic type of object in R is a ***vector***. Vectors are concatenations of data values of the same type. They are defined using the concatenation operator `c()` and are indexed by consecutive integers; subvectors can be retrieved by specifying the indices between square brackets.

```{r}
# numeric vector
c(1, 4, 7)

# character vector
c('blue', 'red')

# indexing
c(1, 4, 7)[1]
c(1, 4, 7)[2]
c(1, 4, 7)[3]
c(1, 4, 7)[2:3]
c(1, 4, 7)[c(1, 3)]
```

Usually objects are assigned names for easy retrieval. Vectors will not show any special object class if the structure is examined; `str()` will simply return the data type, index range, and the values.

```{r}
# assign a name
my_vec <- c(1, 4, 7)

# check structure
str(my_vec)
```

Next up in complexity are ***arrays***. These are blocks of data values of the same type indexed along one or more dimensions. For arrays, `str()` will return the data type, index structure, and data values; when printed directly, data values are arranged according to the indexing.

```{r}
# an array
my_ary <- array(data = c(1, 2, 3, 4, 5, 6, 7, 8), 
           dim = c(2, 4))

my_ary

str(my_ary)

# another array
my_oth_ary <- array(data = c(1, 2, 3, 4, 5, 6, 7, 8), 
           dim = c(2, 2, 2))

my_oth_ary

str(my_oth_ary)
```

For arrays, elements can be retrieved by index coordinates, and slices can be retrieved by leaving index positions blank, which will return all elements along the corresponding indices.

```{r}
# one element
my_ary[1, 2]

# one element
my_oth_ary[1, 2, 1]

# a slice (second row)
my_ary[2, ]

# a slice (first layer)
my_oth_ary[ , , 1]
```

Next there are ***lists***, which are perhaps the most flexible data structure. A list is an indexed collection of any objects.

```{r}
# a list
list('cat', c(1, 4, 7), TRUE)

# a named list
list(animal = 'cat',
     numbers = c(1, 4, 7),
     short = TRUE)
```

List elements can be retrieved by index in double square brackets, or by name.

```{r}
# assign a name
my_lst <- list(animal = 'cat',
               numbers = c(1, 4, 7),
               short = TRUE)

# check structure
str(my_lst)

# retrieve an element
my_lst[[1]]

# equivalent
my_lst$animal
```

Finally, ***data frames*** are type-heterogeneous lists of vectors of equal length. More informally, they are 2D arrays with columns of differing data types. `str()` will essentially show the list structure; but when printed, data frames will appear arranged in a table.

```{r}
# a data frame
my_df <- data.frame(animal = c('cat', 'hare', 'tortoise'),
                    has.fur = c(TRUE, TRUE, FALSE),
                    weight.lbs = c(9.1, 8.2, 22.7))

str(my_df)

my_df
```

The data frame is the standard object type for representing datasets in R. For the most part, modern computing in R is designed around the data frame.

## Packages

R packages are add-ons that can include special functions, datasets, object classes, and the like. They are published software and can be installed using `install.packages('PACKAGE NAME')` and, once installed, loaded via `library('PACKAGE NAME')` or `require('PACKAGE NAME')`.

# Tidyverse basics

We will illustrate the use of tidyverse functions to reproduce the analysis shown in class of the survey data.

::: callout-important
## Action

**Load class survey data**

1.  Open your assigned group's repository project in RStudio. Work in that project so paths are relative to the repository root.
2.  Make sure `background-clean.csv`, `interest-clean.csv`, and `survey-metadata.csv` are in the repository's `data` folder. Your instructor provides these files.
3.  Create a new R script in the repository root and save it as `lab2-tidyverse.R`. Keep the supplied `class-analysis.R` as the starting point for your group's analysis.
4.  Copy the code chunk below to the top of your new script and execute it. If needed, install tidyverse once using `install.packages('tidyverse')` in the Console.
:::

```{r}
library(tidyverse)

# Load the current class data from this repository.
background <- read_csv('data/background-clean.csv', show_col_types = FALSE)
interest <- read_csv('data/interest-clean.csv', show_col_types = FALSE)
metadata <- read_csv('data/survey-metadata.csv', show_col_types = FALSE)

# CSVs store proficiency labels as text; define their order explicitly.
background <- background %>%
  mutate(across(ends_with('.prof'),
                ~ factor(.x, levels = c('beg', 'int', 'adv'), ordered = TRUE)))

nrow(background)
nrow(interest)
```

There are **55 survey responses** overall. The shared background file contains **51 respondents** who consented to share skills and experience; the project-preference file contains **49 respondents** who consented to share that section. All 49 consented to both sections. Each analysis describes the respondents included in the relevant file or join.

The files exclude personal information and free-text narratives. `response_id` is an artificial identifier that connects the two shared sections. Keep these data within the class; do not share outside of the course.

The metadata's `data.file` column identifies the file containing each variable. It also documents `class-summary.csv` and `response-timing.csv`, which contain aggregate summaries used in the lecture slides. Those two files are not required for this lab and cannot be joined to individual responses.

You can view the data in one of two ways:

```{r}
# print the data frame for inspection in the console
background

# In the RStudio Console, run View(background) to open the data viewer.
```

Run `View(metadata)` in the RStudio Console and inspect the datasets. You can also select the metadata entries for one file:

```{r}
metadata %>%
  filter(data.file == 'background-clean.csv') %>%
  select(variable.name, variable.description, variable.type, values.notes)
```

Notice these details before starting:

-   `research` and `domain_specialization` use `Yes`, `No`, and `Unsure`, not numeric 0/1 codes.
-   `upper_div_courses` contains the categories `3-5`, `6-8`, and `9+`, not exact course counts.
-   `courses` stores each respondent's selected courses in one comma-separated string.
-   Proficiency levels are `beg`, `int`, and `adv`. Comfort responses are numeric ratings.

## Concepts

The tidyverse is a collection of packages for data manipulation, visualization, and statistical modeling. Some are specialized, such as `forcats` or `lubridate`, which contain functions for manipulating factors and dates and times, respectively. The packages share some common underlying principles.

1.  Packages are built around the data frame
2.  Functions are designed to work with the pipe operator `%>%`
3.  Packages facilitate readable code

The tidyverse facilitates programming in readable sequences of steps that are performed on data frames. For example:

```{r}
#| eval: false

my_df %>% STEP1() %>% STEP2() %>% STEP3()
```

If it helps, imagine that step 1 is defining a new variable, step 2 is selecting a subset of columns, and step 3 is fitting a model of some kind.

## Tibbles

tidyverse packages use a slight generalization of the data frame called a ***tibble***. For the most part, tibbles behave as data frames do, but they are slightly more flexible in ways you'll encounter later.

For now, think of a tibble as just another name for a data frame.

## The pipe operator `%>%`

In short, `x %>% f(y)` is equivalent to `f(x, y)` .

In other words, the pipe operator 'pipes' the result of the left-hand operation into the first argument of the right-hand function.

```{r}
# a familiar example
my_vec <- c(1, 2, 5) 
str(my_vec)

# use the pipe operator instead
my_vec %>% str()
```

## dplyr verbs

The `dplyr` package contains functions for manipulating data frames (tibbles). The functions are named with verbs that describe common operations.

### Core verbs

::: callout-important
## Action

For each verb listed below, copy the code chunk into your script and execute.

Go through the list with your neighbor and check your understanding by describing what the code example accomplishes.
:::

***filter*** -- filter the rows of a data frame according to a condition and return a subset of rows meeting that condition

```{r}
# filter rows
background %>%
  filter(math.comf > 3)
```

***select*** -- select a subset of columns from a data frame

```{r}
# select a column
background %>%
  select(math.comf)
```

***pull*** -- extract a single column from a data frame

```{r}
# pull a column
background %>%
  pull(research)
```

***mutate*** -- define a new column as a function of existing columns

```{r}
# define a new variable
background %>%
  mutate(avg.comf = (math.comf + prog.comf + stat.comf)/3)
```

These operations can be chained together, for example:

```{r}
# sequence of verbs
background %>%
  filter(stat.prof == 'adv') %>%
  mutate(avg.comf = (math.comf + prog.comf + stat.comf)/3) %>%
  select(avg.comf, research) 
```

<!--# TA COMMENT: take a moment to describe what this sequence did. the goal is for students to get the concept of chaining together operations to find a subset of data and/or compute new quantities -->

::: callout-important
## Action

1.  Write a chain of verbs to find the proficiency ratings of respondents with `research == 'Yes'` and `upper_div_courses == '6-8'`.
2.  Repeat for respondents with `research == 'No'` in the same course-count category. Keep `Unsure` distinct from `No`.
3.  Compare results and discuss with your neighbor: do these suggest any patterns?
:::

### Summaries

***Summaries*** reduce rows to summary values using `summarize()`. For example, among respondents who rated their statistics proficiency as advanced, what proportion reported research experience? `research == 'Yes'` creates a logical vector whose mean is the proportion of `TRUE` values. Here the denominator includes all advanced respondents with a nonmissing research response, including those who answered `Unsure`.

```{r}
# a summary
background %>%
  filter(stat.prof == 'adv') %>%
  mutate(avg.comf = (math.comf + prog.comf + stat.comf)/3) %>%
  select(avg.comf, research) %>%
  summarize(prop.research = mean(research == 'Yes', na.rm = TRUE))

# equivalent
background %>%
  filter(stat.prof == 'adv') %>%
  mutate(avg.comf = (math.comf + prog.comf + stat.comf)/3) %>%
  select(avg.comf, research) %>%
  pull(research) %>%
  {mean(. == 'Yes', na.rm = TRUE)}
```

The advantage of `summarize` , however, is that multiple summaries can be computed at once:

```{r}
background %>%
  filter(stat.prof == 'adv') %>%
  mutate(avg.comf = (math.comf + prog.comf + stat.comf)/3) %>%
  select(avg.comf, research) %>%
  summarize(prop.research = mean(research == 'Yes', na.rm = TRUE),
            med.comf = median(avg.comf, na.rm = TRUE))
```

Use `across()` inside `summarize()` to apply the same summary to several columns. The helper `ends_with()` selects columns with a particular suffix.

```{r}
# average comfort levels across shared background respondents
background %>%
  summarise(across(ends_with('.comf'), ~ mean(.x, na.rm = TRUE)))
```

***Grouped summaries*** are summaries computed separately among subsets of observations. To define a grouping structure using an existing column, use `group_by()` . Notice the 'groups' attribute printed with the output.

```{r}
# create a grouping
background %>%
  group_by(stat.prof)
```

Sometimes it can be helpful to simply count the observations in each group:

```{r}
# count observations
background %>%
  group_by(stat.prof) %>%
  count()
```

To compute a grouped summary, first group the data frame and then specify the summary of interest:

```{r}
# a grouped summary
background %>%
  group_by(stat.prof) %>%
  summarise(across(ends_with('.comf'), ~ mean(.x, na.rm = TRUE)),
            .groups = 'drop')
```

::: callout-important
## Action

**Grouped summaries**

1.  Compute the median comfort level of shared background respondents in each subject area.
2.  Compute the median comfort level of shared background respondents in each subject area after grouping by `upper_div_courses`. Treat this as an ordered category, not an exact numeric count.
3.  Compare and discuss with your neighbor: do you notice any interesting patterns?
:::

## Joining the two survey sections

`inner_join()` matches rows using a shared identifier. Use `response_id`, because the two files have different row counts and cannot be combined by row position.

```{r}
joint <- background %>%
  inner_join(interest, by = 'response_id')

nrow(joint)

joint %>%
  count(type)
```

The join contains 49 respondents who consented to both sections. The `type` codes are `ind` for industry, `lab` for research lab, and `both` for no preference. The last code does not mean that a respondent requested both types of project.

::: callout-important
## Action

Calculate mean programming comfort within each project-preference category. Which respondents are included in this summary? Why is the denominator different from a summary of the full background file?
:::

## tidyr verbs

The `tidyr` package reshapes data frames. We will use `pivot_longer()`, `separate()`, and `pivot_wider()` to make a summary table, then `separate_rows()` to work with course selections.

Suppose we calculate multiple summaries of multiple variables. The initial output has one row and one column for each variable/statistic combination:

```{r}
# many variables, many summaries
comf_sum <- background %>%
  summarise(across(
    ends_with('.comf'),
    list(mean = ~ mean(.x, na.rm = TRUE),
         median = ~ median(.x, na.rm = TRUE),
         min = ~ min(.x, na.rm = TRUE),
         max = ~ max(.x, na.rm = TRUE)),
    .names = '{.col}_{.fn}'
  ))

comf_sum
```

`pivot_longer()` collects the columns into a long table. One column contains the original column name and another contains its value:

```{r}
comf_sum %>%
  pivot_longer(everything(), names_to = 'stat', values_to = 'val')
```

`separate()` splits the original column name at the underscore, giving us the survey variable and the summary statistic:

```{r}
comf_sum %>%
  pivot_longer(everything(), names_to = 'stat', values_to = 'val') %>%
  separate(stat, into = c('variable', 'stat'), sep = '_')
```

`pivot_wider()` then makes one column per summary statistic. The result has three rows, one for each subject, and four summary columns plus the variable-name column:

```{r}
comf_sum %>%
  pivot_longer(everything(), names_to = 'stat', values_to = 'val') %>%
  separate(stat, into = c('variable', 'stat'), sep = '_') %>%
  pivot_wider(names_from = stat, values_from = val)
```

::: callout-important
## Action

Explain what one row represents after each reshaping step. Which step changes the number of rows, and which splits an existing column?
:::

## ggplot2

The `ggplot2` package is for data visualization. The syntax takes some getting used to if you haven't seen it before. We'll just look at one example.

Suppose we want to summarize prior coursework among the shared background respondents. The new data store course selections in a single `courses` column. First use `separate_rows()` to give each selected course its own row:

```{r}
course_responses <- background %>%
  select(response_id, courses) %>%
  separate_rows(courses, sep = ',\\s*') %>%
  mutate(courses = str_trim(courses)) %>%
  filter(!is.na(courses), courses != '') %>%
  distinct(response_id, courses)

course_responses
```

One respondent can now appear on several rows. `distinct()` ensures that a respondent is counted at most once per course.

```{r}
# count respondents selecting each course
classes <- course_responses %>%
  count(courses, name = 'n') %>%
  rename(class = courses) %>%
  mutate(proportion = n / nrow(background))

classes
```

The denominator is **51 shared background respondents**, not 55 total survey responses or the number of rows in `course_responses`. There are no missing course responses in the supplied background file. A course-series selection means at least one course in that series, and courses selected by nobody do not appear in this table. Proportions need not sum to one because respondents selected multiple courses.

We could report the results in a table, in which case perhaps arranging in descending order may be helpful:

```{r}
classes %>% arrange(desc(proportion))
```

Let's say we'd rather plot this data. We'll put the course number on one axis and the proportion of students who took it on the other.

```{r}
# plot it
classes %>%
  ggplot(aes(x = proportion, y = class)) +
  geom_point()
```

These commands work by defining plot layers. In the chunk above, the first argument to `ggplot()` is the data. Then, `aes()` defines an 'aesthetic mapping' of the columns of the input data frame to graphical elements. This defines a set of axes. Then, a layer of points is added to the plot with `geom_point()` ; no arguments are needed because the geometric object ('geom') inherits attributes (x and y coordinates) from the aesthetic mapping.

Again we might prefer to arrange the classes by descending order in proportion.

```{r}
fig <- classes %>%
  ggplot(aes(x = proportion, y = reorder(class, proportion))) +
  geom_point()

fig
```

And perhaps fix the plot labels:

```{r}
# adjust labels
fig + labs(x = 'Proportion of shared background respondents', y = '')
```

Notice that ggplot allows for a plot to be stored by name and then further modified with additional layers.

## Checklist

1.  All actions were completed.
2.  All code chunks were copied into your script.
3.  Your script is saved as `lab2-tidyverse.R` in the repository root, beside the `data` folder.
4.  You can explain why background-only summaries use 51 respondents and analyses joining both sections use 49.
5.  You can explain why course proportions use respondents, rather than course-selection rows, as the denominator.

When extending the group analysis, consult `survey-metadata.csv` before parsing other multiple-selection columns. In particular, `areas` contains option labels with internal commas, such as `Predictive modeling, generally`. Splitting that column at every comma would break the labels. Use complete option labels with `str_detect()` and `fixed()` instead.
