---
title: "Sampling concepts and descriptive analysis"
subtitle: "PSTAT197A Fall 2026"
author: "Dr. Coburn"
institute: 'UCSB'
# bibliography: refs.bib  # Restore if using your course bibliography.
format: 
  revealjs:
    incremental: true
    footer: 'PSTAT197A Fall 2026'
    logo: 'img/ucsbds_hex.png'  # Restore in your course repository.
    fig-width: 4
    fig-height: 2
    fig-align: 'left'
    slide-number: true
execute:
  message: false
  warning: false
  echo: false
  cache: false
---

<!-- ## Lessons from last time -->

<!-- -   add `.DS_Store` to `.gitignore` -->

<!-- -   open repo project in RStudio session (not another project or new session) -->

<!-- -   repo clone directory must be kept intact; can move the entire directory but not individual files -->

<!-- -   use client not terminal, at least to start out -->

<!-- -   others? -->

## Today

-   review sampling concepts

-   introduce class survey data

-   present descriptive analysis

# Sampling concepts

## Samples and populations

***population:*** collection of all subjects/units of interest

***sample:*** subjects/units observed in a study

. . .

statistical methodology strives to account for the possibility that the sample could have been different in order to make reliable inferences about the population based on knowledge of the sampling mechanism

## What if inferences aren't possible?

Even if inference isn't possible, data still have value and could be used for:

-   descriptive analysis of the sample;

-   hypothesis generation;

-   developing analysis pipelines.

## What about prediction?

Prediction is a separate goal but still a form of generalization.

-   samples must reflect a broader population for predictions to be accurate at the population level

<!-- -->

-   if an analyst can't expect sample statistics to provide reliable estimates of population quantities, they shouldn't expect predictions based on the sample to be reliable either

## Common problems

Several issues arise *very* often in practice that compromise or complicate an analyst's ability to make inferences (or predictions). Among them:

-   scope of inference from the sample doesn't match the study population

-   subjects/units are selected haphazardly or by convenience

-   researcher conflates sample size with number of observations, *i.e.,* takes lots of measurements on few subjects/units

## Helpful questions

The following questions can help make an assessment of the scope of inference:

-   (protocol) how were subjects/units chosen for measurement and how were measurements collected?

-   (mechanism) was there any random selection mechanism?

-   (exclusion) are there any subjects/units that couldn't possibly have been chosen?

-   (nonresponse) were any subjects/units selected but not measured?

## Class survey data

-   survey distributed to all students offered enrollment in PSTAT197A fall 2026

-   $n = 55$ responses

-   no random selection

## Can the data support inference?

From a previous year’s reading responses:

> *It depends on the question. If you want to draw conclusions about the pstat197a class specifically, this sample is the population and thus will have reliable data. If you want to draw conclusions about the pstat department as a whole, then this is a bad sample because it is likely biased and thus unreliable*

## Alternative perspectives

The comment points to two ways to view the data:

-   a census of PSTAT197A enrollees

-   a convenience sample of...

    -   capstone applicants OR

    -   students qualified for capstones OR

    -   students interested in data science OR

    -   all UCSB students???

## Is there a right answer?

Either way -- census or convenience sample -- excludes inference.

-   census $\longrightarrow$ no inference needed

-   convenience $\longrightarrow$ no inference possible

. . .

So on a practical level, it won't make much difference for designing an analysis of the survey data.

## Descriptive analysis

Any analysis of survey data should be regarded as *descriptive* in nature:

-   summary statistics and/or models are not reliable measures of any broader population

-   results should be interpreted narrowly in terms of the sample at hand

# Descriptive analysis

## A general approach

**Start simple and add complexity gradually.**

From simpler to more complex consider questions involving:

1.  Sample characteristics
2.  Single-variable summaries
3.  Multivariate summaries
4.  Model-based outputs (estimates, predictions, etc.)

## Questions of interest

**Sample characteristics**

-   How do respondents describe their gender identity?

**Single-variable summaries**

-   Among the students offered a seat in PSTAT197, what fields of study are the students most interested in?

-   What level of comfort do students interested in data analysis at UCSB have with mathematics?

## Questions of interest

**Multivariate summaries**

-   Are students who ranked themselves as strong in statistics, mathematics, and computing more likely or less likely to select an 'industry' project as the project type that they want to work on?

**Model-based outputs**

-   Are there distinct groups of students in the class defined by self-assessed proficiencies and/or comfort levels with mathematics, statistics, and programming?

## Sample characteristics

Aggregate summaries of all 55 respondents; personal information is not linked to shared records.

```{r}
library(tidyverse)
library(broom)
library(knitr)
background <- read_csv('data/background-clean.csv', show_col_types = FALSE)
interest <- read_csv('data/interest-clean.csv', show_col_types = FALSE)
class_summary <- read_csv('data/class-summary.csv', show_col_types = FALSE)
timing <- read_csv('data/response-timing.csv', show_col_types = FALSE)
summary_table <- function(which_group) {
  class_summary %>% filter(group == which_group) %>%
    select(category, n, percent)
}
tbl1 <- summary_table('standing')
tbl2 <- summary_table('gender')
tbl3 <- summary_table('race')
tbl0 <- summary_table('consent') %>%
  separate(category, into = c('Skills consent', 'Project consent'), sep = ' / ') %>%
  select(-percent) %>% pivot_wider(names_from = 'Project consent', values_from = n)
```

::: panel-tabset
### Class standing

```{r}
kable(tbl1,
      col.names = c("Class Standing", "Count", "Percent"))
```

### Gender

```{r}
kable(tbl2,
      col.names = c("Gender Identity", "Count", "Percent"))
```

### Race

```{r}
kable(tbl3,
      col.names = c("Race / Ethnicity", "Count", "Percent"))
```

### Data sharing

**Columns**: consent to share project preferences

**Rows**: consent to share background and preparation

```{r}
kable(tbl0)
```
:::

## Majors

::: panel-tabset
### Primary major

```{r, fig.width=7, fig.height=3.5}
class_summary %>% filter(group == 'primary_major') %>%
  ggplot(aes(x = n, y = forcats::fct_reorder(category, n))) +
  geom_col(fill = '#003660') + geom_text(aes(label = n), hjust = -0.2) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = 'Respondents', y = NULL) + theme_minimal()
```

### Second major

```{r, fig.width=7, fig.height=3.5}
class_summary %>% filter(group == 'second_major', category != 'None reported') %>%
  ggplot(aes(x = n, y = forcats::fct_reorder(category, n))) +
  geom_col(fill = '#003660') + geom_text(aes(label = n), hjust = -0.2) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = 'Respondents reporting a second major', y = NULL) + theme_minimal()
```
:::

## Response timing

```{r, fig.width=7, fig.height=3.5}
timing %>%
  mutate(day = as.POSIXct(paste(date, '12:00:00'), tz = 'America/Los_Angeles')) %>%
  ggplot(aes(day, n)) +
  geom_col(width = 20*60*60, fill = '#003660') +
  geom_vline(xintercept = as.numeric(as.POSIXct('2026-10-04 23:59:00',
                 tz = 'America/Los_Angeles')), color = '#d1495b', linetype = 2) +
  scale_x_datetime(date_breaks = '2 days', date_labels = '%b %d',
                   timezone = 'America/Los_Angeles') +
  labs(x = 'Submission date (Pacific time)', y = 'Responses') + theme_minimal()
```

Deadline: **October 4, 11:59 p.m. Pacific** (dashed line). **54 on time; 1 late.**

## Shared data

-   51 respondents consented to share skills and experience.
-   49 respondents consented to share project preferences.
-   49 consented to both; use these respondents for joint analyses.
-   Consent applies separately to each section. Summaries describe the corresponding subset.

## Privacy

The following information has been removed from the dataset distributed to the class:

-   personal information from section 1 of the survey

-   long text and free response answers, contain some personal details

-   each section’s responses when the student did not consent to share that section

-   individual timestamps; only daily response counts are provided

-   type distinction between research experiences

## Single-variable summaries

What level of comfort do students interested in data analysis at UCSB have with mathematics?

::: panel-tabset
### **Comfort**

```{r}
background %>%
  select(contains('comf')) %>%
  summarise_all(.funs = list(mean = mean, 
                             median = median,
                             min = min, 
                             max = max)) %>%
  gather(stat, val) %>%
  separate(stat, into = c('variable', 'stat'), sep = '_') %>%
  spread(stat, val) %>%
  kable()
```

### Proficiency (numeric)

```{r}


background %>%
  select(contains(".prof")) %>%
  # Normalize text & collapse longer phrases into beg/int/adv
  mutate(across(
    everything(),
    ~ {
      x <- str_to_lower(str_trim(.x))
      case_when(
        str_detect(x, "adv") ~ "adv",
        str_detect(x, "inter") ~ "int",
        str_detect(x, "beg|little") ~ "beg",
        TRUE ~ NA_character_
      )
    }
  )) %>%
  mutate(across(
    everything(),
    ~ factor(.x, levels = c("beg", "int", "adv"), ordered = TRUE)
  )) %>%
  rename_with(~ gsub("\\.prof", "", .x)) %>%
  mutate(across(everything(), as.numeric)) %>%
  summarise(across(
    everything(),
    list(mean = ~ round(mean(.x, na.rm = TRUE), 2),
         median = ~ median(.x, na.rm = TRUE)),
    .names = "{.col}_{.fn}"
  )) %>%
  pivot_longer(everything(),
               names_to = c("variable", "stat"),
               names_sep = "_",
               values_to = "value") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  kable()

```

### Proficiency (factor)

```{r}
# background %>%
#   select(contains(".prof")) %>%
#   mutate(across(everything(), ~ factor(.x, levels = c("beg", "int", "adv")))) %>%
#   rename_with(~ gsub("\\.prof", "", .x)) %>%
#   pivot_longer(everything(), names_to = "skill", values_to = "level") %>%
#   count(skill, level) %>%
#   pivot_wider(names_from = skill, values_from = n, values_fill = 0) %>%
#   kable()

background %>%
  select(contains(".prof")) %>%
  # Normalize text and classify correctly
  mutate(across(
    everything(),
    ~ {
      x <- str_to_lower(str_trim(as.character(.x)))
      case_when(
        str_detect(x, "adv") ~ "adv",
        str_detect(x, "int|intermediate") ~ "int",
        str_detect(x, "beg|little") ~ "beg",
        TRUE ~ NA_character_
      )
    }
  )) %>%
  rename_with(~ gsub("\\.prof", "", .x)) %>%
  pivot_longer(everything(), names_to = "skill", values_to = "level") %>%
  count(level, skill, name = "n") %>%
  filter(!is.na(level)) %>%  # drop unmatched NA rows
  mutate(level = factor(level, levels = c("beg", "int", "adv"))) %>%
  complete(level, skill, fill = list(n = 0)) %>%
  arrange(level) %>%
  pivot_wider(names_from = skill, values_from = n, values_fill = 0) %>%
  kable()
```
:::

## Multivariable summaries {.scrollable}

Are students who ranked themselves as strong in statistics, mathematics, and computing more likely or less likely to select an 'industry' project as the project type that they want to work on?

::: panel-tabset
### Counts

Proficiency: beginner = 1, intermediate = 2, advanced = 3. Fixed score intervals keep tied scores together.

```{r}
tbl <- interest %>%
  inner_join(background, by = 'response_id') %>%
  select(type, contains('prof')) %>%
  mutate(across(contains('prof'), 
                ~as.numeric(factor(.x, levels = c('beg', 
                                       'int', 
                                       'adv'))))) %>%
  rename_with(~gsub('.prof', '', .x)) %>%
  rowwise() %>%
  mutate(mean.proficiency = mean(c(math, stat, prog))) %>%
  ungroup() %>%
  mutate(mean.proficiency.fac = cut(mean.proficiency,
         breaks = c(-Inf, 2, 2.5, Inf),
         labels = c('1.00–2.00', '>2.00–2.50', '>2.50–3.00'))) %>%
  select(type, mean.proficiency.fac) %>%
  count(mean.proficiency.fac, type) %>%
  complete(mean.proficiency.fac, type = c('ind', 'lab', 'both'), fill = list(n = 0)) %>%
  pivot_wider(names_from = type, values_from = n, values_fill = 0)

tbl %>% kable()
```

### **Proportions**

```{r}
tbl %>%
  rowwise() %>%
  mutate(n = sum(ind, lab, both)) %>%
  ungroup() %>%
  mutate(across(c(ind, lab, both), ~round(.x/n, 3))) %>%
  kable()
```
:::

## Combinations {.scrollable}

Consider the distinct *combinations* of comfort and proficiency ratings (separately):

::: panel-tabset
### Proficiency

```{r}
# unique combinations of proficiency ratings
proficiency <- background %>%
  select(contains('.prof')) %>%
  mutate_all(~factor(.x, levels = c('beg', 'int', 'adv'))) %>%
  mutate_all(as.numeric)

proficiency %>% 
  rename_with(~gsub('.prof', '', .x)) %>%
  group_by(prog, math, stat) %>%
  count() %>%
  kable()
```

### Comfort

```{r}
# unique combinations of comfort ratings
comfort <- background %>%
  select(contains('comf'))

comfort %>% 
  rename_with(~gsub('.comf', '', .x)) %>%
  group_by(prog, math, stat) %>%
  count() %>%
  kable()
```
:::

## Clustering

Can students be grouped based on combinations of preferences and comfort levels?

::: panel-tabset
### Centers

```{r}
# cluster responses into three groups
set.seed(92922)
clust <- bind_cols(proficiency, comfort) %>%
  kmeans(centers = 3, nstart = 25)

clust %>%
  broom::tidy() %>%
  select(-withinss) %>%
  kable(digits = 3)
```

### Visualization

```{r, fig.width = 5, fig.height = 4, fig.align='center'}
# Center the same variables used for k-means, then project with PCA.
cluster_matrix <- bind_cols(proficiency, comfort)
projection <- prcomp(cluster_matrix, center = TRUE, scale. = FALSE)
as_tibble(projection$x[, 1:2]) %>%
  mutate(cluster = factor(clust$cluster)) %>%
  ggplot(aes(PC1, PC2, color = cluster)) +
  geom_point(size = 2, alpha = 0.75) +
  labs(x = 'Principal component 1', y = 'Principal component 2', color = 'Cluster')
```

### Method

Clustering method, "k means", groups data by nearest Euclidean distance to each of $k$ centers. $k$ is user-specified; the method finds the centers that minimize within-cluster variance.

### Interpretation

Cluster labels depend on this year's fitted centers. The summaries below update with the data.

```{r, results='asis'}
centers <- as_tibble(clust$centers) %>%
  mutate(cluster = row_number(),
         mean_proficiency = rowMeans(across(ends_with('.prof'))),
         mean_comfort = rowMeans(across(ends_with('.comf'))))
for (i in seq_len(nrow(centers))) {
  cat(sprintf('- Cluster %d (n = %d): mean proficiency %.2f; mean comfort %.2f.\n\n',
      i, clust$size[i], centers$mean_proficiency[i], centers$mean_comfort[i]))
}
```

Proficiency uses scores 1–3; comfort uses the observed numeric ratings. K-means uses unscaled variables, so their scales affect the result. The clusters are descriptive.
:::

## Assignment

Your task is to extend this analysis with your group by the assigned deadline.

. . .

Here are some ideas:

-   explore variable associations further (*e.g.,* coursework and self-evaluations)

-   experiment with clustering on different variable subsets or using different methods

-   summarize domain or area of interest variables (requires some text manipulation)

## Next time

Most of next meeting we'll devote to planning your group's task.

-   Do a little brainstorming on your own

-   Come with a few questions/ideas
